Parallel Solving of Multiple Information-Coordinated Global Optimization Problems

### 2021-03-06 2020/JPDC_MCO/2020_JPDC.tex
JPDC правки

The first and obvious approach is to use local methods that allow one to efficiently solve optimization problems with unimodal objective functions \cite{c1}. In the case of multiextremal problems, local methods should be applied repeatedly to find global minima using different initial points. Research in this direction began as early as the 1960s and is still relevant today \cite{c10,c11,c12}. Unfortunately, the problem of choosing initial points still remains open -- even using a significant number of initial points does not guarantee a reliable solution of global optimization problems. Another obvious approach is to construct in the search domain  a number of grids in whose nodes values of functions to be minimized are computed \cite{c2,c3,c4,c5}. To reduce computational complexity of uniform grids, this approach envisages different methods for construction of non-uniform grids, which would be denser only in the vicinity of the global minimum. This line of research is developing in the field of so-called nature-inspired metaheuristic methods, including, for example, genetic algorithms, bee algorithms, particle swarm optimization, simulated annealing, ant colony optimization, and firefly algorithms, etc. \cite{c4,c13,c14,c15}. A large number of applied problems have been solved using these methods, however, as before, it is also difficult to obtain guaranteed estimates of the accuracy when finding globally-optimal solutions for these algorithms.

@@

The paper proposes a further complication in formulating the decision making problems, when not only one, but several global optimization problems need to be solved simultaneously. Such problems are encountered, for example, when one needs to find several efficient solutions of multicriteria optimization problems or when part of the variable parameters in the optimization problem can take only a certain number of discrete values. For the first time, the formulation of such multiple global optimization (MGO was considered in \cite{c25,c26}, and then was investigated in \cite{c27,c28}. The relatively limited attention given to MGO problems is most likely due to the high computational complexity involved for solving even one GO problem.

In general, MGO problems can be solved by a simple sequential (or parallel) solution of individual problems from a set. However, this approach results in a many-fold increase in the computational complexity of solving such problems. Meanwhile, MGO problems may have a certain informational interrelation -- a similar connectivity is observed, for example, when methods based on scalarization of the vector efficiency criterion are applied to solve multicriteria optimization (MCO) problems. In this case, it becomes necessary to solve multiple scalar (in the general case, global) optimization problems in order to find several efficient decisions  for MCO problems -- see, for example, \cite{c29,c30,c31,c32,c33}. Such interconnectedness of MGO problems can be used to reduce significantly the amount of computation performed when solving the entire set of problems.

The approach proposed in this paper assumes that the computational values of any function of the set of MGO problems can be recalculated to the values of any other functions of the set without any time consuming calculations. In such cases, all the search information obtained in solving a particular problem can be used in solving all other problems in the set. Such repeated reuse of search information can significantly reduce the amount of computation performed, up to just a few iterations when solving the next MGO problems. In addition, the presence of informational connectivity significantly expands the possibilities of a parallel solution for MGO problems. The paper proposes two new parallel computational schemes: a cooperative scheme, which means the search information between the problems is exchanged when solving several MGO problems in parallel, and a competitive scheme, which occurs when MGO problems solved in parallel compete with each other for the use of available computational resources.

@@

As a result, if the values of efficiency criteria $g_i(y)$, $1 \leq i \leq s$, are computed at some point $y=\bar{y}$ of the search domain $D$, then the value of the scalar criterion $G(\lambda,y)$ can be computed at the point $\bar{y} \in D$ for any values of the coefficients $\lambda \in \Lambda$. In this case the transformation of $Z$ from (\ref{eq:5}) is determined by the expression (\ref{eq:7}) at fixed values of the criteria $g_l (y)$,$1 \leq l \leq m$) \cite{c27,c28}.

The ability to quickly convert the calculated values of optimized functions determines the need to accumulate and use search information obtained in the calculation process
\begin{equation}\label{eq:9}
\omega_k (j)=\{(y^i,f^i=f_j(y^i)):1 \leq i \leq k \},1 \leq j \leq s,
\end{equation}
where $j$, $1 \leq j \leq s$, is the number of the current function being minimized $f_j (y)$ of the set $F$ from (\ref{eq:3}), k is the number of iterations of the global search, $y^i \in D$, $1 \leq i \leq k$,  are points in which the values of the optimized problems of the set $F$ were computed, and $f^i$, $1 \leq i \leq k$, are the values of the function $f_j (y)$ at these points (the procedure for calculating the value of the function will be called a trial). Global optimization methods can use this search information, thereby ensuring the ability to solve each successive problem of the $F$ set, taking into account the results of all previous calculations. Numerical experiments that were performed (see Section \ref{sec:5}), confirm that such repeated use of search information reduces the number of iterations of the global search required to solve the next problem of the set $F$ (even to the point of performing just a few search iterations).

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{fig1}
  \caption{Examples of approximations of the Peano space-filling curves (the approximation density is equal to 2, 3 and 4 correspondingly) \cite{c5}}
  \label{fig:1}
\end{figure}

@@

For the multidimensional generalizations of the algorithms proposed in \cite{c34,c35}, the characteristic is an estimate of the minimum possible value of the function being minimized on the interval, i.e.
\begin{equation}\label{eq:17}
R(i)=0.5 H \rho_i - 0.5 (z_{i-1} + z_i), \rho_i=(x_i - x_{i-1})^{1/N}, 1 < i \leq k,
\end{equation}
where $H$ the values $z_i$, $1<i \leq k$, are from $A_k$, H is the H\"older constant from (\ref{eq:12}) for the reduced global optimization problem to be solved and $N$ is the dimensionality of the problem from (\ref{eq:1}). For the multidimensional generalized global search algorithm (GSA) \cite{c6,c23}, developed in the framework of the information-statistical approach, the characteristic has the form
\begin{equation}\label{eq:18}
R(i)=m \rho_i+\frac{(z_i-z_{i-1})^2}{m \rho_i }-2(z_{i-1}+z_i ), \rho_i=(x_i-x_{i-1})^{1/N}  ,1 < i \leq k,
\end{equation}
where $m$ is the numerical estimate of the H\"older constant obtained from the available search information $A_k$ from (\ref{eq:13})

@@

A detailed description of the GSA algorithm is given in Algorithm \ref{alg:1}(see also \cite{c6}). The inputs of the algorithm are the function to be minimized $f(y)$\footnote{ }, the search domain $D$ from (\ref{eq:2}), the reliability parameter $r > 1$, the required accuracy $\varepsilon > 0$ and the maximum permissible number of global search iterations $kmax$. The output of the algorithm is an estimate of the global minimum point $y_k^* =y(x_k^*)$, the value of the function to be minimized at this point $z_k^* =f(y_k^*)$  and the number of performed global search iterations $k$. In the course of calculations, the search information $A_k$\footnote{ } from (\ref{eq:13}) is used, which contains the points of previously performed global search iterations $x_i$, $1 \leq i \leq k$, ordered in ascending order, and the values of the function to be minimized $z_i=f(y(x_i))$, $1\leq i \leq k$.

@@

An obvious way to apply parallel computing is to solve the problems of the set F in parallel. Undoubtedly this approach is straightforward and does not require any additional efforts to be  implemented. In this case, if a sufficient number of computational devices (cores / processors) is available, the time needed to solve all problems from the F set will be determined by the time used in solving the most time-consuming problem. However, this time may turn out to be quite long and necessitate the use of parallel computing to solve even one global optimization problem. However, well-known and widely used parallelization methods are poorly applicable in the case of global optimization problems. For example, the search domain distribution between processors for global optimization problems leads to the global minimum of the problems to be solved being located on only one of the processors, and all other processors will perform redundant calculations.

@@

To implement this approach, it is necessary to ensure that at each iteration of the global optimization problem, not one, but several points in the search domain $D$ are selected for parallel calculation of the values of the functions to be minimized at these points. In keeping with the fact that characteristically represented global search algorithms use the characteristics of the intervals $R(i)$, $1 < i \leq k$, from (\ref{eq:15}), to assess the possibility of locating the global minimum in the subinterval of the interval [0,1], it is most reasonable to select the points of the next iteration of the global search in the intervals with the maximum values of the characteristics \cite{c6}.

@@

@@

Assume, as before, that $q>1$ is the number of available computational devices (cores/processors) with shared memory, and $ns>1$ is the number of optimization problems\footnote{Without lost of generality, we will use sequential numbering of problems solved in parallel from 1 to $ns$.}  solved in parallel. Each optimization problem will compete for available computational resources -- the distribution of resources between problems can be carried out, as before, based on the values of interval characteristics. To calculate the characteristics, the problems solved in parallel have their own separate sets of search information $A_k (is)$, $1 \leq is \leq ns$, from (\ref{eq:13}), which match the set of iteration points $x_i$, $1 \leq i \leq k$, of the global search, but differ, accordingly, in the values $z_i (is)$, $1 \leq is \leq k$, of the minimized functions $f_is (y)$, $1 \leq is \leq ns$, from the set $F(y)$. As a result, in this case, to specify the intervals where the characteristics are computed, two indices $( is, t )$ are required, where $is$, $1 \leq is \leq ns$, is the number of the function to be minimized  $f_is (y)$ and  $t$, $1 \leq t \leq k$, is the number of the interval in the search information $A_k (is)$.

It should also be noted that when implementing this approach to compare the interval characteristics in different global optimization problems, the values of the function being minimized in the available search information should be normalized\footnote{Without loss of generality, we will use sequential numbering of problems solved in parallel from 1 to $ns$.}
\begin{equation}\label{eq:25}
z(is)_i'=\frac{z(is)_i - z(is)_k^*}{m(is)_k}, 1 \leq i \leq k,1 \leq is \leq ns,
\end{equation}
where $z(is)_k$, $1 \leq is \leq ns$, are the estimates (\ref{eq:23}) of the minimum values of the functions being minimized, and $m(is)_k$, $1 \leq is \leq ns$, are the estimates (\ref{eq:19}) of the H\"older constants $H_{is}$, $1 \leq is \leq ns$, from (\ref{eq:12}). Further, when performing the steps of the algorithm, the converted values $z(is)_i'$, $1 \leq i \leq k$, $1 \leq is \leq ns$, should be used when calculating the characteristics of intervals and points of new iterations of the global search.

@@

\begin{algorithm}[]
\caption{PSEUDO CODE OF THE PMGSA-2 ALGORITHM} \label{alg:3}
\scriptsize
\hspace*{\algorithmicindent} \textbf{Input}: $F(y)$, $D$, $r$, $\varepsilon$, $kmax$, $ns$;\\
\hspace*{\algorithmicindent} \textbf{Output}: $Y$, $Z$, $k$;
\begin{algorithmic}[1]
\State Initialize $ k=2, A_2=\{(x_1=0,y_1=y(x_1),z_1=f(y(x_1)),(x_2=1,y_2=y(x_2),z_2=f(y(x_2))\} $;
\State Set $stop = false$;
\While{$k < kmax$ \textbf{and not} $stop$}
  \ForEach { $is = 1..ns$ }
	  \State Calculate the estimate of the H\"older constant $m$ from (\ref{eq:19}) using $A_k$;
		\State Find the estimate of the global minimum value $z(is)_k^*$ of the function $f_{is}(y)$ in accordance with (\ref{eq:23});
		\State Set $A_k'(is)=A_k (is)$;
		\State Normalize the values $z(is)_i'$, $1 \leq i \leq k$, in $A_k'(is)$ in accordance with (\ref{eq:25});
		\State Set $m_k (is)=1$;
	\EndFor

	\State Set $T = \{ \emptyset \}, XK = \{ \emptyset \}$;
	\ForEach { $i = 1..q$ } 
	// q is the number of available cores/processors
	  \State Find the interval with the maximum characteristic 
		\begin{equation*}
			R_{is} (t)= \argmax_{1 \leq is \leq ns, 1 < i \leq k} R(i),i \notin T 
	  \end{equation*}
		using $A_k' (is)$;
		\State Calculate the iteration point $x^{k+i}$ using $A_k' (is)$ in accordance with (\ref{eq:21});
    \State Insert $t$ in $T$, is in $IS$ and $x^{k+i}$ in $XK$;

@@

Let us briefly explain the rules for performing a parallel iteration of the global search. The characteristics of the intervals are calculated for all ns problems to be solved in parallel, and q intervals with the maximum values of the characteristics are selected among them (lines 12-16). As a result, the available computational resources are distributed dynamically among problems -- computational devices can be distributed evenly between problems, or they can be allocated to only one problem with the best interval characteristics. After calculating the values of the functions being minimized, the new search data obtained is added to the search information of all solved problems; regardless of where the intervals for iterating the global search were selected, the search information of all solved problems is updated with a full set of new search data (lines 19-24).

The maximum efficiency of the proposed PMGSA-2 algorithm is achieved if the global minimum in parallel problems is reached at the same point in the search domain. In any case, getting additional search information from simultaneously solved problems can significantly reduce the time needed to solve each global search problem. This conclusion is rather obvious if the global minima points of the functions $f_i (y)$, $1 \leq i \leq s$, of the set $F$ from (\ref{eq:3}) are situated in the same small neighborhood of the search domain $D$ from (\ref{eq:2}). But even in the case when the global minima points of different functions differ much from each other, the values of different functions $f_i (y)$, $1 \leq i \leq s$, of the set $F$ computed in the process of global search can be used for each function separately. Thus, in PMGSA-2 the execution of one parallel iteration using $q$ computational units adds $q$ computation results. Increasing the amount of available search information allows adaptive global search algorithms to select points for the next iterations of the global search in a more reasonable manner. In the limiting case, when performing sufficiently many $iterations \gg 1$, the amount of search information can be sufficient to find the global minimum point of the next problem in the set $F$ without any additional iterations. This effect is proved also by the results of numerical experiments in Section \ref{sec:5}. An example of how PMGSA-2 selects the intervals with the maximum characteristics is given in Fig.\ref{fig:3}.

@@

When using computational devices with shared memory, the number of cores / processors is quite limited -- the number of cores on one computational node usually does not exceed 12-24. For increasing possible parallelism, it becomes inevitable to use a distributed memory computer with several computational nodes.

Let $p>1$ be the number of computational nodes with distributed memory, each of which has $q>1$ computational cores (that is, the total number of cores in a computational system is $p*q$).

@@

2) Each computational node is applied for solving q optimization problems using the PMGSA-2 algorithm.

@@

In this way, in keeping with the above rules, the PMGSA-3 algorithm is distributed -- the problems of the set $F(y)$ are solved independently on different computational nodes. During the global search process, computational nodes exchange the calculation results that are obtained; after each iteration of the global search, the search information for each problem being solved is updated with $q$ calculated values of the function being minimized. Thus, the more computational nodes there are (and accordingly, the problems that are solved simultaneously), the faster, as a rule, each optimization problem is solved. This is evidenced, in particular, by the results of numerical experiments that were performed (see Section \ref{sec:5}). An example of how PMGSA-3 performs an optimization iteration is given in Fig. \ref{fig:4}.

@@

Numerical experiments were performed on the supercomputers Lobachevsky (University of Nizhni Novgorod) and Endeavour (Intel Cluster). The first supercomputer was used for software development. The numerical results were obtained by the second one with the following computational nodes: 2 Intel Xeon Platinum 8260L, 2.4 GHz, 256 GB RAM (i.e. a total of 48 CPU cores were available on each node). The executable program code was built by using the Intel Parallel Studio XE 2019 software package. The numerical experiments were performed using the Globalizer system \cite{c36}. This system offers an opportunity to solve computationally intensive multicriteria multiextremal optimization problems using the parallel global search algorithms presented in the paper.

@@

\begin{table}[ht]
\centering
\caption{Comparison of the effectiveness of multicriteria optimization algorithms \cite{c37})}
\label{tab:1}
\begin{tabular}{cccccc}
\hline
Method                                                                                     & MC    & SEMO  & NUC   & BLO   & \textbf{MGSA} \\ \hline
Iteration number                                          & 500   & 500   & 515   & 498   & \textbf{370}   \\
\begin{tabular}[c]{@{}c@{}}Number of Pareto optimal points\end{tabular} & 67    & 104   & 29    & 68    & \textbf{100}    \\
HV                                                        & 0.300 & 0.312 & 0.306 & 0.308 & \textbf{0.316} \\
DU                                                        & 1.277 & 1.116 & 0.210 & 0.175 & \textbf{0.101} \\ \hline
\end{tabular}
\end{table}

These numerical results show that the MGSA algorithm has a noticeable advantage in almost all performance indicators when compared with the methods of multicriteria optimization being considered, even when solving relatively simple MGO problems.

@@

The results of the numerical experiments are presented in Table \ref{tab:2}. The first three rows of the Table contain information about the computational resources used during the experiments. The first row marked as ``p'' indicates the number of computational nodes used. The number of cores used within a single node is presented in the second row marked as ``q''. The total number of cores used is presented in the third row marked as ``p*q''. The rows four through six show the average number of iterations performed by PMGSA-1,2,3 for solving a single MCO problem accordingly.

@@

The results of the numerical experiments demonstrate that all methods have found the sufficient approximation of the Pareto domain. The lowest number of iterations performed the PMGSA-2 method. In Fig. \ref{fig:6}, the computed approximation of the Pareto domain obtained by the PMGSA-3method is presented.

@@

The availability of informational connectivity significantly expands the possibilities of parallel solutions of the entire set of multiple global optimization problems. The paper proposes two parallel computational schemes: a cooperative scheme, when several problems are solved simultaneously, with the search information that is obtained being exchanged between problems, and a competitive scheme, when problems that are simultaneously being solved compete with each other for using the available computational resources.

>>>>

The first and obvious approach is to use local methods that allow one to efficiently solve optimization problems with unimodal objective functions \cite{c1}. In the case of multiextremal problems, local methods should be applied repeatedly to find global minima using different initial points. Research in this direction began as early as the 1960s and is still relevant today \cite{c10,c11,c12}. Unfortunately, the problem of choosing initial points still remains open -- even using a significant number of initial points does not guarantee a reliable solution of global optimization problems. Another obvious approach is to construct a grid in the search domain, at the nodes of which the values of functions to be minimized are computed \cite{c2,c3,c4,c5}. To reduce computational complexity of uniform grids, this approach envisages different methods for construction of non-uniform grids, which would be denser only in the vicinity of the global minimum. This line of research is developing in the field of so-called nature-inspired metaheuristic methods, including, for example, genetic algorithms, bee algorithms, particle swarm optimization, simulated annealing, ant colony optimization, and firefly algorithms, etc. \cite{c4,c13,c14,c15}. A large number of applied problems have been solved using these methods, however, as before, it is also difficult to obtain guaranteed estimates of the accuracy when finding globally-optimal solutions for these algorithms.

@@

The paper proposes a further complication in formulating the decision making problems, when not only one, but several global optimization problems need to be solved simultaneously. Such problems are encountered, for example, when one needs to find several efficient solutions of multicriteria optimization problems or when part of the variable parameters in the optimization problem can take only a certain number of discrete values. For the first time, the formulation of such multiple global optimization (MGO) was considered in \cite{c25,c26}, and then was investigated in \cite{c27,c28}. The relatively limited attention given to MGO problems is most likely due to the high computational complexity involved for solving even one GO problem.

In general, MGO problems can be solved by a simple sequential (or parallel) solution of individual problems from a set. However, this approach results in a many-fold increase in the computational complexity of solving such problems. Meanwhile, individual GO problems that are part of the same MGO problem may have a certain informational interrelation -- a similar connectivity is observed, for example, when methods based on scalarization of the vector efficiency criterion are applied to solve multicriteria optimization (MCO) problems. In this case, it becomes necessary to solve multiple scalar (in the general case, global) optimization problems in order to find several efficient decisions  for MCO problems -- see, for example, \cite{c29,c30,c31,c32,c33}. Such interconnectedness of GO problems can be used to reduce significantly the amount of computation performed when solving the entire set of problems.

The approach proposed in this paper assumes that the computational values of any function of the same MGO problem can be recalculated to the values of any other functions of the set without any time consuming calculations. In such cases, all the search information obtained in solving a particular problem can be used in solving all other problems in the set. Such repeated reuse of search information can significantly reduce the amount of computation performed, up to just a few iterations when solving the next GO problems. In addition, the presence of informational connectivity significantly expands the possibilities of a parallel solution for MGO problems. The paper proposes two new parallel computational schemes: a cooperative scheme, which means the search information between the problems is exchanged between computational devices with distributed memory used for solving several GO problems in parallel, and a competitive scheme, which occurs when GO problems solved in parallel compete with each other for the use of available computational resources with shared memory.

@@

As a result, if the values of efficiency criteria $g_i(y)$, $1 \leq i \leq m$, are computed at some point $y=\bar{y}$ of the search domain $D$, then the value of the scalar criterion $G(\lambda,y)$ can be computed at the point $\bar{y} \in D$ for any values of the coefficients $\lambda \in \Lambda$. In this case the transformation of $Z$ from (\ref{eq:5}) is determined by the expression (\ref{eq:7}) at fixed values of the criteria $g_l (y)$,$1 \leq l \leq m$) \cite{c27,c28}.

The ability to quickly convert the calculated values of optimized functions determines the need to accumulate and use search information obtained in the calculation process
\begin{equation}\label{eq:9}
\omega_k (j)=\{(y^i,f^i=f_j(y^i)):1 \leq i \leq k \},1 \leq j \leq s,
\end{equation}
where $j$, $1 \leq j \leq s$, is the number of the current function being minimized $f_j (y)$ of the set $F$ from (\ref{eq:3}), $k$ is the number of iterations of the global search, $y^i \in D$, $1 \leq i \leq k$,  are points in which the values of the optimized problems of the set $F$ were computed, and $f^i$, $1 \leq i \leq k$, are the values of the function $f_j (y)$ at these points (the procedure for calculating the value of the function will be called a trial). Global optimization methods can use this search information, thereby ensuring the ability to solve each successive problem of the $F$ set, taking into account the results of all previous calculations. Numerical experiments that were performed (see Section \ref{sec:5}), confirm that such repeated use of search information reduces the number of iterations of the global search required to solve the next problem of the set $F$ (even to the point of performing just a few search iterations).

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{fig1}
  \caption{Examples of approximations of the Peano space-filling curves (the approximation density is equal to 2, 3 and 4 correspondingly) \cite{c6}}
  \label{fig:1}
\end{figure}

@@

For the multidimensional generalizations of the algorithms proposed in \cite{c34,c35}, the characteristic is an estimate of the minimum possible value of the function being minimized on the interval, i.e.
\begin{equation}\label{eq:17}
R(i)=0.5 H \rho_i - 0.5 (z_{i-1} + z_i), \rho_i=(x_i - x_{i-1})^{1/N}, 1 < i \leq k,
\end{equation}
where the values $z_i$, $1<i \leq k$, are from $A_k$, $H$ is the H\"older constant from (\ref{eq:12}) for the reduced global optimization problem to be solved and $N$ is the dimensionality of the problem from (\ref{eq:1}). For the multidimensional generalized global search algorithm (GSA) \cite{c6,c23}, developed in the framework of the information-statistical approach, the characteristic has the form
\begin{equation}\label{eq:18}
R(i)=m \rho_i+\frac{(z_i-z_{i-1})^2}{m \rho_i }-2(z_{i-1}+z_i ), \rho_i=(x_i-x_{i-1})^{1/N}  ,1 < i \leq k,
\end{equation}
where $m$ is the numerical estimate of the H\"older constant obtained from the available search information $A_k$ from (\ref{eq:13})

@@

A detailed description of the GSA algorithm is given in Algorithm \ref{alg:1} (see also \cite{c6}). The inputs of the algorithm are the function to be minimized $f(y)$\footnote{ }, the search domain $D$ from (\ref{eq:2}), the reliability parameter $r > 1$, the required accuracy $\varepsilon > 0$ and the maximum permissible number of global search iterations $kmax$. The output of the algorithm is an estimate of the global minimum point $y_k^* =y(x_k^*)$, the value of the function to be minimized at this point $z_k^* =f(y_k^*)$  and the number of performed global search iterations $k$. In the course of calculations, the search information $A_k$\footnote{ } from (\ref{eq:13}) is used, which contains the points of previously performed global search iterations $x_i$, $1 \leq i \leq k$, ordered in ascending order, and the values of the function to be minimized $z_i=f(y(x_i))$, $1\leq i \leq k$.

@@

An obvious way to apply parallel computing is to solve the problems of the set $F$ in parallel. Undoubtedly this approach is straightforward and does not require any additional efforts to be  implemented. In this case, if a sufficient number of computational devices ({cores / processors}) is available, the time needed to solve all problems from the set $F$ will be determined by the time used in solving the most time-consuming problem. However, this time may turn out to be quite long and necessitate the use of parallel computing to solve even one global optimization problem. However, well-known and widely used parallelization methods are poorly applicable in the case of global optimization problems. For example, the search domain distribution between processors for global optimization problems leads to the global minimum of the problems to be solved being located on only one of the processors, and all other processors will perform redundant calculations.

@@

To implement this approach, it is necessary to ensure that at each iteration of the global optimization problem, not one, but several points in the search domain $D$ are selected for parallel calculation of the values of the functions to be minimized at these points. In keeping with the fact that characteristically represented global search algorithms use the characteristics of the intervals $R(i)$, $1 < i \leq k$, from (\ref{eq:15}), to assess the possibility of locating the global minimum in the subintervals of the interval [0,1], it is most reasonable to select the points of the next iteration of the global search in the intervals with the maximum values of the characteristics \cite{c6}.

@@

Input data of this algorithm are the set of functions $F(y)$ from (\ref{eq:3}), search domain $D$ from (\ref{eq:2}), reliability parameter $r>1$, required accuracy $\varepsilon > 0$, maximum possible number of calculations of the values of the functions to be minimized $kmax$ and the number of available computational devices ({cores/processors}) with shared memory $q>1$. The results of executing the algorithm are the sets of estimates of global minimum points of the functions being minimized $Y$ and values of the functions $Z$ in these points, and the total number of executed iterations of global search $k$. As before, in the course of calculations we use the search information $A_k$, which contains the points of earlier iterations of global search $x_i$, $1 \leq i \leq k$, ordered in ascending order, and values of the current function being minimized $z_i=f(y(x_i))$, $1 \leq i \leq k$.

@@

Assume, as before, that $q>1$ is the number of available computational devices (cores/processors) with shared memory, and $ns>1$ is the number of optimization problems\footnote{Without lost of generality, we will use sequential numbering of problems solved in parallel from 1 to $ns$.}  solved in parallel. Each optimization problem will compete for available computational resources -- the distribution of resources between problems can be carried out, as before, based on the values of interval characteristics. To calculate the characteristics, the problems solved in parallel have their own separate sets of search information $A_k (is)$, $1 \leq is \leq ns$, from (\ref{eq:13}), which match the set of iteration points $x_i$, $1 \leq i \leq k$, of the global search, but differ, accordingly, in the values $z_i (is)$, $1 \leq i \leq k$, of the minimized functions $f_{is} (y)$, $1 \leq is \leq ns$, from the set $F(y)$. As a result, in this case, to specify the intervals where the characteristics are computed, two indices $( is, t )$ are required, where $is$, $1 \leq is \leq ns$, is the number of the function to be minimized  $f_{is}(y)$ and  $t$, $1 \leq t \leq k$, is the number of the interval in the search information $A_k (is)$.

It should also be noted that when implementing this approach to compare the interval characteristics in different global optimization problems, the values of the function being minimized in the available search information should be normalized\footnote{After transformation (\ref{eq:25}), $z_k^*=0$ and $m_k=1$ are carried out for all $ns$ parallel problems.}
\begin{equation}\label{eq:25}
z(is)_i'=\frac{z(is)_i - z(is)_k^*}{m(is)_k}, 1 \leq i \leq k,1 \leq is \leq ns,
\end{equation}
where $z(is)_k$, $1 \leq is \leq ns$, are the estimates (\ref{eq:23}) of the minimum values of the functions being minimized, and $m(is)_k$, $1 \leq is \leq ns$, are the estimates (\ref{eq:19}) of the H\"older constants $H_{is}$, $1 \leq is \leq ns$, from (\ref{eq:12}). Further, when performing the steps of the algorithm, the converted values $z(is)_i'$, $1 \leq i \leq k$, $1 \leq is \leq ns$, should be used when calculating the characteristics of intervals and points of new iterations of the global search.

@@

\begin{algorithm}[]
\caption{PSEUDO CODE OF THE PMGSA-2 ALGORITHM} \label{alg:3}
\scriptsize
\hspace*{\algorithmicindent} \textbf{Input}: $F(y)$, $D$, $r$, $\varepsilon$, $kmax$, $ns$;\\
\hspace*{\algorithmicindent} \textbf{Output}: $Y$, $Z$, $k$;
\begin{algorithmic}[1]
\State Initialize $ k=2, A_2=\{(x_1=0,y_1=y(x_1),z_1=f(y(x_1)),(x_2=1,y_2=y(x_2),z_2=f(y(x_2))\} $;
\State Set $stop = false$;
\While{$k < kmax$ \textbf{and not} $stop$}
  \ForEach { $is = 1..ns$ }
	  \State Calculate the estimate of the H\"older constant $m(is)_k$ from (\ref{eq:19}) using $A_k(is)$;
		\State Find the estimate of the global minimum value $z(is)_k^*$ of the function $f_{is}(y)$ in accordance with (\ref{eq:23});
		\State Set $A_k'(is)=A_k (is)$;
		\State Normalize the values $z(is)_i'$, $1 \leq i \leq k$, in $A_k'(is)$ in accordance with (\ref{eq:25});
		\State Set $m_k (is)=1$;
	\EndFor

	\State Set $T = \{ \emptyset \}, XK = \{ \emptyset \}$;
	\ForEach { $i = 1..q$ } 
	// q is the number of available cores/processors
	  \State Find the interval with the maximum characteristic using $A_k' (is)$
		\begin{equation*}
			R_{is} (t)= \argmax_{1 \leq is \leq ns, 1 < i \leq k} R(i),i \notin T;
	  \end{equation*}

		\State Calculate the iteration point $x^{k+i}$ using $A_k' (is)$ in accordance with (\ref{eq:21});
    \State Insert $t$ in $T$, $is$ in $IS$ and $x^{k+i}$ in $XK$;

@@

Let us briefly explain the rules for performing a parallel iteration of the global search. The characteristics of the intervals are calculated for all $ns$ problems to be solved in parallel, and $q$ intervals with the maximum values of the characteristics are selected among them (lines 12-16). As a result, the available computational resources are distributed dynamically among problems -- computational devices can be distributed evenly between different problems, or they can be allocated to only one problem with the best interval characteristics. After calculating the values of the functions being minimized, the new search data obtained is added to the search information of all solved problems; regardless of where the intervals for iterating the global search were selected, the search information of all solved problems is updated with a full set of new search data (lines 19-24).

The maximum efficiency of the proposed PMGSA-2 algorithm is achieved if the global minimum in parallel problems is reached at the same point in the search domain. In any case, getting additional search information from simultaneously solved problems can significantly reduce the time needed to solve each global search problem. This conclusion is rather obvious if the global minima points of the functions $f_i (y)$, $1 \leq is \leq ns$, of the set $F$ from (\ref{eq:3}) are situated in the same small neighborhood of the search domain $D$ from (\ref{eq:2}). But even in the case when the global minima points of different functions differ much from each other, the values of different functions $f_i (y)$, $1 \leq is \leq ns$, of the set $F$ computed in the process of global search can be used for each function separately. Thus, in PMGSA-2 the execution of one parallel iteration using $q$ computational devices adds $q$ computation results. Increasing the amount of available search information allows adaptive global search algorithms to select points for the next iterations of the global search in a more reasonable manner. In the limiting case, when performing sufficiently many iterations $k \gg 1$, the amount of search information can be sufficient to find the global minimum point of the next problem in the set $F$ without any additional iterations. This effect is proved also by the results of numerical experiments in Section \ref{sec:5}. An example of how PMGSA-2 selects the intervals with the maximum characteristics is given in Fig.\ref{fig:3}.

@@

When using computational devices with shared memory, the number of {cores/processors} is quite limited -- the number of cores on one computational node usually does not exceed 12-24. For increasing possible parallelism, it becomes inevitable to use a distributed memory computational system with several computational nodes.

Let $p>1$ be the number of computational devices with distributed memory, each of which has $q>1$ computational cores (that is, the total number of cores in a computational system is $p*q$).

@@

2) Each computational node is applied for solving $q$ optimization problems using the PMGSA-2 algorithm.

@@

In this way, in keeping with the above rules, the PMGSA-3 algorithm is distributed -- the problems of the set $F(y)$ are solved independently on different computational nodes. During the global search process, computational nodes exchange the calculation results that are obtained; after each iteration of the global search, the search information for each problem being solved is updated with $p*q$ calculated values of the function being minimized. Thus, the more computational nodes there are (and accordingly, the problems that are solved simultaneously), the faster, as a rule, each optimization problem is solved. This is evidenced, in particular, by the results of numerical experiments that were performed (see Section \ref{sec:5}). An example of how PMGSA-3 performs an optimization iteration is given in Fig. \ref{fig:4}.

@@

Numerical experiments were performed on the supercomputers Lobachevsky (University of Nizhni Novgorod) and Endeavour (Intel Cluster). The first supercomputer was used for optimization software system. The numerical results were obtained by the second one with the following computational nodes: 2 Intel Xeon Platinum 8260L, 2.4 GHz, 256 GB RAM (i.e. a total of 48 CPU cores were available on each node). The executable program code was built by using the Intel Parallel Studio XE 2019 software package. The numerical experiments were performed using the Globalizer system \cite{c36}. This system offers an opportunity to solve computationally intensive multicriteria multiextremal optimization problems using the parallel global search algorithms presented in the paper.

@@

\begin{table}[ht]
\centering
\caption{Comparison of the effectiveness of multicriteria optimization algorithms)}
\label{tab:1}
\begin{tabular}{cccccc}
\hline
Method                                                                                     & MC    & SEMO  & NUC   & BLO   & \textbf{MGSA} \\ \hline
Iteration number                                          & 500   & 500   & 515   & 498   & \textbf{370}   \\
\begin{tabular}[c]{@{}c@{}}Number of Pareto optimal points\end{tabular} & 67    & 104   & 29    & 68    & \textbf{100}    \\
HV                                                        & 0.300 & 0.312 & 0.306 & 0.308 & \textbf{0.316} \\
DU                                                        & 1.277 & 1.116 & 0.210 & 0.175 & \textbf{0.101} \\ \hline
\end{tabular}
\end{table}

These numerical results show that the MGSA algorithm has a noticeable advantage in all performance indicators when compared with the methods of multicriteria optimization being considered, even when solving relatively simple MGO problems.

@@

The results of the numerical experiments are presented in Table \ref{tab:2}. The first three rows of the Table contain information about the computational resources used during the experiments. The first row marked as ``$p$'' indicates the number of computational nodes used. The number of cores used within a single node is presented in the second row marked as ``$q$''. The total number of cores used is presented in the third row marked as ``$p*q$''. The rows four through six show the average number of iterations performed by PMGSA-1,2,3 for solving a single MCO problem accordingly.

@@

The results of the numerical experiments demonstrate that all methods have found the sufficient approximation of the Pareto domain. The lowest number of iterations performed the PMGSA-3 method. In Fig. \ref{fig:6}, the computed approximation of the Pareto domain obtained by the PMGSA-2 method is presented.

@@

The availability of informational connectivity significantly expands the possibilities of parallel solutions of the entire set of multiple global optimization problems. The paper proposes two parallel computational schemes: a cooperative scheme, which means the search information between the problems is exchanged between computational devices with distributed memory used for solving several GO problems in parallel, and a competitive scheme, which occurs when GO problems solved in parallel compete with each other for the use of available computational resources with shared memory..
